Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-08 TACO-paper-v2/coffee.tex
Section 3: copy editing

As already emphasized, the variations in the structure of
mathematical expressions and in loop trip counts, although typically
limited to the order of tens of iterations, render the optimization process
challenging, requiring distinct sets of transformations to bring performance
closest to the machine peak in different problems. For example, the Burgers problem in Listing~\ref{code:burgers},
given the large number of arrays accessed, suffers from high register pressure,
whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
not; this intuitively suggests that the two problems require a
different treatment, based on an in-depth analysis of both data and
iteration spaces. Furthermore, domain-knowledge enables
transformations that a general-purpose compiler could not apply,
making the optimization space even larger. In this context, our goal
is to understand the relationship between distinct code
transformations, their impact on cross-loop arithmetic intensity, and to what
extent their composability is effective in a class of problems and
architectures.

@@

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then, two
dimensional arrays are padded by rounding the number of columns to the
nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the original size of a
basis function array is 3$\times$3 and $\mbox{\texttt{VL}}=4$ (e.g. AVX processor,
with 256-bit long vector registers and 64-bit double-precision
floats). In this case, a padded version of the array will have size
3$\times$4. The compiler is explicitly told about data alignment
using suitable pragmas; for example, in the case of the Intel compiler, the annotation \texttt{$\#$pragma vector aligned} is added over a loop
(as shown in later figures) to inform that all of the memory accesses in the loop body will be properly aligned. This allows the compiler to issue aligned load and store instructions, which are notably faster than unaligned ones.

@@

We work around these limitations with source-level loop-invariant code
motion. In particular, we pre-compute all values that an invariant
sub-expression assumes along its fastest varying dimension. This is
implemented by introducing a temporary array per invariant
sub-expression and by adding a new loop to the nest. At the price of
extra memory for storing temporaries, the gain is that lifted terms
can be auto-vectorized as part of an inner loop. Given the short trip
counts of our loops, it is important to achieve auto-vectorization of
hoisted terms in order to minimize the percentage of scalar
instructions, which could otherwise be significant. It is also worth
noting that, in some problems (e.g. Helmholtz), invariant
sub-expressions along \texttt{j} are identical to those along
\texttt{k}, and both loops iterate over the same iteration space, as
anticipated in Section~\ref{sec:background}. In these cases, we safely
avoid redundant pre-computation.

@@

\KwSty{void} helmholtz(\KwSty{double} A[4][4], \KwSty{double} **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~// Do generalized loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~\KwSty{$\_\_$m256d} r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz
  problem after application of \emph{op-vect} on top of the
  optimizations shown in Listing~\ref{code:helmholtz-licm}. In this
  example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

>>>>

As already emphasized, the variations in the structure of
mathematical expressions and in loop trip counts, although typically
limited to the order of tens of iterations, render the optimization process
challenging, requiring distinct sets of transformations to bring performance
closest to the machine peak in different problems. For example, the Burgers problem in Listing~\ref{code:burgers},
given the large number of arrays accessed, suffers from high register pressure,
whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
not; this intuitively suggests that the two problems require a
different treatment, based on an in-depth analysis of both data and
iteration spaces. Furthermore, domain knowledge enables
transformations that a general-purpose compiler could not apply,
making the optimization space even larger. In this context, our goal
is to understand the relationship between distinct code
transformations, their impact on cross-loop arithmetic intensity, and to what
extent their composability is effective in a class of problems and
architectures.

@@

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then,
two dimensional arrays are padded by rounding the number of columns to
the nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the
original size of a basis function array is 3$\times$3 and
$\mbox{\texttt{VL}}=4$ (e.g. AVX processor, with 256-bit long vector
registers and 64-bit double-precision floats). In this case, a padded
version of the array will have size 3$\times$4. The compiler is
explicitly told about data alignment using suitable pragmas; for
example, in the case of the Intel compiler, the annotation
\texttt{$\#$pragma vector aligned} is added before the loop (as shown in
later figures) to inform that all of the memory accesses in the loop
body will be properly aligned. This allows the compiler to issue aligned
load and store instructions, which are notably faster than unaligned
ones.

@@

We work around these limitations with source-level loop-invariant code
motion. In particular, we pre-compute all values that an invariant
sub-expression assumes along its fastest varying dimension. This is
implemented by introducing a temporary array per invariant
sub-expression and by adding a new loop to the nest. At the price of
extra memory for storing temporaries, the gain is that lifted terms
can be auto-vectorized as part of an inner loop. Given the short trip
counts of our loops, it is important to achieve auto-vectorization of
hoisted terms in order to minimize the percentage of scalar
instructions, which could otherwise be significant. It is also worth
noting that, in some problems, for instance Helmholtz, invariant
sub-expressions along \texttt{j} are identical to those along
\texttt{k}, and both loops iterate over the same iteration space, as
anticipated in Section~\ref{sec:background}. In these cases, we safely
avoid redundant pre-computation.

@@

\KwSty{void} helmholtz(\KwSty{double} A[4][4], \KwSty{double} **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~// Do generalized loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~~~// \emph{load} and \emph{set} intrinsics \\
~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
~~~~~~~~// One \emph{permute\_pd} intrinsic per \texttt{k}-loop \emph{load}\\
~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
~~~~~~~~// One \emph{permute2f128\_pd} intrinsic per \texttt{k}-loop \emph{load}\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~\KwSty{$\_\_$m256d} r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz
  problem after application of \emph{op-vect} on top of the
  optimizations shown in Listing~\ref{code:helmholtz-licm}. In this
  example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}
